---
title: "Live_coding"
author: "Julia"
date: "2023-06-08"
output: html_document
---

Class 5
Lesson 05 cont'd

Working with data tables
```{r}
library(data.table) #load library data.table
data(iris) #load data iris into our session

dt_iris <- as.data.table(iris) #convert iris to a data table
setkey(dt_iris, Species) #set key as species 

dt_iris

dt_iris[,sum((Petal.Length + Petal.Width >2))] #Counting the number of cases where Petal.Length plus Petal.Width is greater than 2

dt_iris[Species == "setosa", sum((Petal.Length + Petal.Width > 2))] #Counting the number of cases where Species is “setosa” and Petal.Length plus Petal.Width is greater than 2

dt_iris[Species == "setosa" , .N] #Counting observations in current group

table(dt_iris$Species) #Can use table() function to count how many species of each group there are

dt_iris[, .N, by = (Species)] #alternative way to count how many species of each group

dt_iris[Petal.Width > 1,
        .(mean(Petal.Length)), keyby = .(Species)] #select only rows with petal length > 1 , calculate mean petal length for selected rows for each species

dt_iris[order(Species, -Petal.Width)] #Ordering by Species and then by reverse Petal.Width:

dt_iris[, .N, .(Petal.Length > 4, Petal.Width >1)] #Grouping those that have Petal.Length greater/not than 4, and Petal.Width greater/not than 1

dt_iris[, lapply(.SD, mean), by = Species] #taking the mean of all columns (except the grouping column, Species) using lapply and .SD

dt_iris[, head(.SD, 4), by = Species] #getting the first four rows for all columns, for each Species
```


Exercise
1. Tidy the data below:
```{r}
data <- tibble(
  group = c("treat", "control"),
  survival = c(17, 11),
  deceased = c(3, 9)
)

data

data_longer <- pivot_longer(data, cols = c("survival", "deceased"), names_to = "sur/deceased", values_to= "counts")

pivot_wider(data_longer, names_from = "sur/deceased", values_from = "counts")
```
2. Join the dataset flights to the dataset airlines. What should the key(s) be? What do the different types of joins look like?
```{r}
library(nycflights13)
data("flights")
data("airlines")

dim(flights)
dim(airlines)

airlines_2 <- airlines[-1,] #removed row 1 to demonstrate antijoin subsetting

flights %>%
  inner_join(airlines, by ="carrier")

flights %>%
  left_join(airlines, by = "carrier")

flights  %>%
  anti_join(airlines_2, by = "carrier")
```

3. Using flights and data.table, group based on cases that have dep_delay < 0 and those that have arr_delay > 0 and count the number in each group. 
How many groups/rows are there? How many in each group?
```{r}
flights_dt <- as.data.table(flights) 
flights_dt[, .N, .(dep_delay < 0, arr_delay > 0)]
```

Lesson 06

Functions
examples of functions:
```{r}

##function that calculates the standard error
standard_error_mean <- function(numeric_vector){
  SEM <- sd(numeric_vector) /  sqrt(length(numeric_vector)-1)
  return(SEM)
}


x <- seq(1,10)
standard_error_mean(x)

##function that prints a string
print_string <-
  function(character){
    print(character)
  }

print_string("Hello, my name is Julia")


#function that replaces setosa with NA in the iris dataset
rename_species <- function(data_set){
  ifelse(data_set$Species == "setosa", NA, data_set$Species)
}

rename_species(iris)
```


For Loops
```{r}
for (num in 1:10) {
  print(num*5)
}

nrow(iris)
iris$Sepal.Length[1] + iris$Sepal.Length[2]
iris$Sepal.Length[2] + iris$Sepal.Length[3]


for (i in 2:nrow(iris)){
  iris$previous_combo[i] <-
    iris$Sepal.Length[i] + 
    iris$Sepal.Length[i-1]
}
iris
```

Ways to loop
```{r}
letters

for (letter in letters){
  print(letter)
}

for (num in 1:3){
  print(now() + num)
}

outputs <- c()

for (i in 1:5){
  outputs <- c(outputs, i)
}
outputs
```

while loops
```{r}
i = 1

while (i <= 10){
  print(i)
  print(i * 5)
  i = i + 1
}
```

if/else
```{r}
grades <- tibble(grade = c(94,87,73,60))
grades %>%
  mutate(letter = case_when(
    grade >= 80 ~ "A",
    grade >= 70 ~ "B",
    grade >= 60 ~ "C",
    TRUE ~ "F" #If all other conditions are false
  ))
```

```{r}
iris %>%
  map_dbl(mean)

iris %>%
  map_chr(typeof)

iris %>%
  map(summary)
```

```{r}
x <- list(1,1,1)
y <- list(10,20,30)

x
y

map2(x,y, ~.x + .y)
```

Simulation
```{r}

hist(runif(1000)) #uniform distribution

runif(5)
runif(5)

set.seed(123)
runif(5)

set.seed(100)
runif(5)

runif(10, min = 1, max = 20)


hist(rnorm(1000)) #normal distribution

set.seed(123)
rnorm(10, mean = 5, sd = 1)


set.seed(103)
sample(c("a","b","c"),
       size = 10,
       replace = TRUE)

set.seed(103)
sample(c("a","b","c"),
       size = 10,
       replace = TRUE)


set.seed(123)

simulated_data <- tibble(
  X = runif(10,min = 0, max = 20),
  Y = 3 *X + rnorm(10, mean = 0, sd = 1)
)

simulated_data

simulated_data %>%
  ggplot(aes( x = X, y = Y)) + geom_point()
```
