---
title: "class_2"
author: "Julia"
date: "2023-06-01"
output: html_document
---
Class 2
Lesson 02

```{r}
sessionInfo() # to see base packages in R
library(help = "base")
library(help = "stats")
library(help = "datasets")
data() # access built in functions in R
```

To download a package from CRAN
```{r}
#for example
install.packages("tidyverse") # to download
library("tidyverse") #to attach the package to the current R session
?tidyverse
ls("package:tidyverse") # to list all functions
tidyverse_packages()

```

What to do if you come across an issue?
```{r}
log("a")
```
RStudio Community to get help
https://community.rstudio.com/

reprex: reproducible examples
1. Environment; 
```{r}
sessionInfo()
```

2. Toy dataset if you're working with a large data frame, subset!

3. Code
```{r}
#install.package("reprex)
library(reprex)

reprex({
  ####paste code
})

reprex({
  log("a")
})
```

Lesson 03

Different element types

```{r}
exampleElement <- "Hello word"
typeof(exampleElement) # "character"
```
```{r}
1 # is treated like 1.0
typeof(1) # "double"
```
```{r}
1L # 1
typeof(1L) # "integer"
```

Logical type: TRUE and FALSE are reserved words denoting logical constants
```{r}
TRUE
typeof(TRUE) # "logical"
```

complex consists of a real part and an imaginary part, 
```{r}
typeof(3 + 1i) # "complex"
```

atomic vectors
c function: Combine Values into a Vector or List
```{r}
c()
?c
```

numeric vector
```{r}
doubleVector <- c(1, 2, 3, 4)
typeof(doubleVector) # "double"
```
```{r}
integerVector <- c(1L, 2L, 3L, 4L)
typeof(integerVector) # "integer"
integerVector
length(integerVector)
```

Note: Data that are of type double can store general numeric values. 
An integer can only take whole number values and requires an L to be 
added after the number to tell R you want the number stored as an integer

logical vectors
```{r}
logicalVector <- c(TRUE, FALSE, TRUE, TRUE)
typeof(logicalVector) # "logical"
```

The sum() function will treat TRUE as 1 and FALSE as 0
```{r}
logicalVector2 <- c(TRUE, FALSE, TRUE, TRUE, NA)
typeof(logicalVector2)
sum(logicalVector) #3 TRUES
sum(logicalVector2, na.rm = TRUE) # 3 TRUES
?sum
```

Character vectors
```{r}
characterVector <- c("hello", "world", "DSI")
typeof(characterVector)
```

Coercion
converting one type to another of vector
1. Automatic - R decides
2. Explicit - User decides

Automatic /implicit coercion
The general principle is that R tries to preserve as much information as possible during implicit coercion. 

```{r}
mixVector <- c("hello", TRUE, 1L, 2.5, 4+1i)
mixVector
typeof(mixVector) # "character"
```
```{r}
mixVector2 <- c(TRUE, 1L, 2.5, 4+1i)
typeof(mixVector2) # "complex"
```

Explicit coercion - as.* function

```{r}
characterVector2 <- c("1", "2", "3")
typeof(characterVector2) # "character"
```

```{r}
numericVector2 <- as.numeric(characterVector2)
typeof(numericVector2) # "double"
mean(numericVector2)
```

```{r}
numericVector3 <- as.integer(characterVector2)
typeof(numericVector3) # "integer"
mean(numericVector3)
```

```{r}
numericVector4 <- as.logical(characterVector2)
typeof(numericVector4) # "logical"
numericVector4 # NA NA NA

```

vector recycled
```{r}
1:5 + 6:10
```
# 1 2 3 4 5
# 6 7 8 9 10

```{r}
1:2 + 1:4
```
# 1 2 1 2
# 1 2 3 4

```{r}
1:2  + 1:5
```
# 1 2 1 2 1
# 1 2 3 4 5


Naming
you can use the names() function to assign names to the elements of a vector or retrieve the names assigned to a vector
```{r}
?names
exampleVector <- c(1:5)
exampleVector
names(exampleVector) # NULL
length(exampleVector)
# letters is reserved for alphabet letters
names(exampleVector) <- letters[1:5]
names(exampleVector) # "a" "b" "c" "d" "e"
exampleVector

namedVector <- c(a = 100, b = 200, c = 80)
namedVector

namedVector['c']
namedVector[c('a','b')]
namedVector[3]
namedVector[c(-1,-2)]

```

```{r}
namedVector[c(TRUE, TRUE, FALSE)]
namedVector[namedVector == 100]
```


```{r}
mylist <- list(7, "abc", FALSE)
mylist
str(mylist)

mylist <- list(a = 1:4, b = "zyx", c = list(-1, -5))
mylist[[2]]
mylist$b
```

```{r}
mytibble <- tibble(x = 1:5,
       y = 1,
       z = x ^ 2 + y)
mytibble
```
```{r}
library(tidyverse)
library(dplyr)
data("iris")
head(iris)

iris_tibble<- as_tibble(iris)
iris_tibble
```

```{r}
iris$Species

iris[["Species"]]

iris %>% 
  .$Species

iris %>%
  .[["Species"]]
```

```{r}
library(stringr)
str_length("this is a string")

str_c("this is a string", "6" ,sep = " ")

str_sub("this is a string", 1, 4)

str_to_lower("HI MY NAME IS JULIA")
str_to_upper("hi my name is julia")
str_to_title("hi my name is julia")

myfruits <- c("apple", "banana",
              "clementime", "dragonfruit")

?str_view

str_view(myfruits, "an")

str_view(myfruits, ".a")

mystring <- c(".", "x","y")
str_view(mystring, "\\.")

mystring <- c("Hello ! \\")
str_view(mystring, "\\\\")

str_view(myfruits, "^a")
str_view(myfruits, "a$")

mystring <- "abcccdeee"
str_view(mystring, "cc?")
str_view(mystring, "cc+")
str_view(mystring, "c[de]+")

mystring <- c("abab", "cdcd", "efgh")
str_view(mystring, "(..)\\1", match = T)

mystring <- c("banana", "dodo", "apple")
str_detect(mystring, "(..)\\1")
str_subset(mystring, "(..)\\1")
str_count(mystring, "(..)\\1")
str_locate(mystring, "(..)\\1")
```