---
title: "Live Coding"
author: "Julia Gallucci"
date: "2023-12-11"
output: html_document
---
Class 1
Lesson 00,01,02


Get working directory
```{r}
getwd() # get working directory
```

Set working directory (NOT recommended method)
```{r}
# setwd("/Users/as/Desktop/DSI-IntroR") 
```
Set working directory (Recommended method)

- Session -> Set working directory -> choose directory
- Or use files tab


A variable/ R object
As mentioned, the top right quadrant is where variables you made are stored. lets make a variable!
```{r}
exampleVariable <- 1 # <- is the assignment operator
exampleVariable = 1 # not recommended
exampleVariable # autoprinting
```

you can remove variables so they no longer appear in your environment
```{r}
rm(exampleVariable) 
```

You can comment outside of chunks, or within chunks using the #
```{r}
# This is a new comment. 
```

R is case sensitive
```{r}
exampleVariable <- 1
examplevariable
exampleVariable
exampleVariable2 <- 2

rm(exampleVariable, exampleVariable2)
```

You can also overwrite variables or 'reassign' them 
```{r}
exampleVariable <- 1
exampleVariable
exampleVariable <- 10 # dynamic variable typing
exampleVariable
exampleVariable <- "a"
exampleVariable
```


Variable convention
you can call your variable in the console by simply typing in the variable name

Object names camelCase styling
```{r}
exampleVariable
```
Object names snake_case styling
```{r}
example_variable
```
Object names dot.style not recommended
```{r}
example.variable
```
R has reserved words
```{r}
NA # not available
NaN # not a number
TRUE
FALSE
T
F
# don't use these to create objects
```

Write and run your own variable!
```{r}
my_name <- "Julia"
todays_date <- "December 11, 2023"
```

Mathematical operations
+  add
-  subtract
*  multiplication
/  division

Let’s pretend we want to write some code, that performs basic math operations. 
I.e., we want to code 2+2
```{r}
2 + 2 
```
Write and run your own basic addition in your script! 
We can do something similar, directly in the console (type 2 + 2 in the console!)
```{r}
```

Built-in functions; originally a statistical language
```{r}
mean()
sqrt()
seq()
```
Structure of a function
functionName(argumentName) 
args, return values
this gives return values

```{r}
nums <- seq(1,3)
mean(nums)
```

Help documentation
example from help documentation
```{r}
?mean
help(mean)

```


In-class exercises
Lesson 01 slide 43

1. load our libraries
```{r}
library(tidyverse)
library(faraway)
```

2. load the datatset "broccoli"
```{r}
data("broccoli")
broccoli
```


In-class exercises
Lesson 02 

1. Create toy dataset
```{r}
toy_data <- data.frame(ID = 1:4,
                       name = c("Abe", "Becca", "Calvin", "Danica"),
                                age = c(27,32,63,55), membership = c("yes","no","yes","yes"))

toy_data
```

2. Create a reprex
```{r}
#install.packages("reprex")
library(reprex)

reprex({
  toy_data$Name
})
```
# End of class 11 December











