---
title: "in-class coding"
author: "Julia Gallucci"
date: "18/12/2023"
output: html_document
---

Load in our libraries
```{r}
library(ggplot2)
library(tidyverse)
```

Load in our dataset
```{r}
CES_data <- read_csv("CES_data_short.csv")
```
Overall syntax
```{r}
#ggplot(data = dataset, mapping = aes(x = variable1, y = variable2))

#or

# dataset %>%
# ggplot(aes(x = variable1, y = variable2)) + layer + another layer...
```

Barplot
```{r}
table(CES_data$cps19_votechoice)

CES_data %>%
  ggplot(aes(x = cps19_votechoice)) +
  geom_bar()

?geom_bar
```
Histogram
```{r}
CES_data %>%
  mutate(age = 90 - cps19_yob) %>%
  ggplot(aes(x = age)) + geom_histogram()
```
Scatterplot
```{r}
CES_data %>%
  mutate(age = 90 - cps19_yob) %>%
  ggplot(aes(x = age, y = cps19_household)) + geom_point()
```

Exercise
1. Make a barplot
```{r}
CES_data %>%
  ggplot(aes(x = cps19_province)) + geom_bar()
```
2. Make a histogram
```{r}
table(CES_data$cps19_party_rating_24)

CES_data %>%
  ggplot(aes(x = cps19_party_rating_24)) +
  geom_histogram()
```
3. Make a scatterplot
```{r}
str(CES_data)

CES_data %>%
  ggplot(aes(x=cps19_yob, y=cps19_party_rating_23)) + geom_point()
```
Customization

Add labels
```{r}
CES_data %>%
  ggplot(aes(x = cps19_gender)) + geom_bar() + labs(x = "Self-identified gender", y = "Counts", title = "Gender frequencies in the CES data")
```
Fill bars by a variable
```{r}
CES_data %>%
  ggplot(aes(x = cps19_age, fill = cps19_gender)) +
  geom_histogram()
```
Fill bars by a color
```{r}
CES_data %>%
  ggplot(aes(x = cps19_age)) +
  geom_histogram(fill = "orange")
```
color lines by a variable
```{r}
CES_data %>%
  ggplot(aes(x = cps19_age,
             y = cps19_household,
             color = cps19_gender)) +
  geom_smooth()
```
color lines by a specified color
```{r}
CES_data %>%
  ggplot(aes(x = cps19_age,
             y = cps19_household)) +
  geom_smooth(color = "blue")
```
Change size of datapoints by a third variable
```{r}
CES_data %>%
  ggplot(aes(x = cps19_age,
             y = cps19_income_number,
             size = cps19_household)) +
  geom_point()
```
Change size of datapoints by a specified size number
```{r}
CES_data %>%
  ggplot(aes(x = cps19_age,
             y = cps19_income_number)) +
  geom_point(size = 5)
```
Use two geoms with a shares aesthetic 
```{r}
CES_data %>%
  ggplot(aes(x = cps19_age,
             y = cps19_income_number)) +
  geom_point() +
  geom_smooth()
```
Use two geoms, each with their own aesthetic
```{r}
CES_data %>%
  ggplot() +
  geom_point(aes(x = cps19_age,
                 y = cps19_income_number)) +
  geom_line(aes(x = cps19_age, y = mean(cps19_income_number, na.rm = TRUE)))
```
Facets - wrap by a categorical variable
```{r}
CES_data %>%
  ggplot(aes(x = cps19_age)) +
  geom_histogram() +
  facet_wrap(facets = "cps19_gender")
```
Change the theme of the overall plot
```{r}
CES_data %>%
  ggplot(aes(x = cps19_age)) +
  geom_histogram() +
  facet_wrap(facets = "cps19_gender") +
  theme_classic()
```
Add discrete x labels to reduce clutter
```{r}
CES_data %>%
  ggplot(aes(x = cps19_votechoice)) +
  geom_bar() + scale_x_discrete(labels = c(
    "Liberal Party" = "Liberal",
    "Conservative Party" = "Conservative",
    "Bloc Québécois" = "Bloc",
    "Green Party" = "Green",
    "People's Party" = "PPC",
    "Another party" = "Other",
    "Don't know/ Prefer not to answer"= "DK/PNTA"
  ))
```

Fill color by a variable
specify ordering of x labels
change position to dodge so bars are side-by-side rather than stacked
```{r}
CES_data %>%
  ggplot() +
  geom_bar(aes(x = factor(cps19_votechoice,
                          levels = c("Liberal Party", "Conservative Party", "Bloc Québécois","Green Party", "People's Party", "Another party", "Don't know/ Prefer not to answer")), fill = cps19_gender), position = "dodge", width = 0.5) + scale_x_discrete(labels = c(
    "Liberal Party" = "Liberal",
    "Conservative Party" = "Conservative",
    "Bloc Québécois" = "Bloc",
    "Green Party" = "Green",
    "People's Party" = "PPC",
    "Another party" = "Other",
    "Don't know/ Prefer not to answer"= "DK/PNTA"
  ))
```

wrap histogram of conservative party rating by thoughts on education spending
```{r}
CES_data %>%
  ggplot() +
  geom_histogram(aes(x = cps19_party_rating_24), bins = 10) + facet_wrap(facets = "cps19_spend_educ", nrow = 1)
```
level facet to provide an order that makes more sense
```{r fig.height=5, fig.width=5}
CES_data %>%
  ggplot() +
  geom_histogram(aes(x = cps19_party_rating_24), bins = 10) + facet_wrap(~factor(cps19_spend_educ, levels = c("Spend less","Spend about the same as now", "Spend more", "Don't know/ Prefer not to answer")), nrow = 1) + labs(x = "Conservative party rating", y = "Counts", title = "Views on education spending and the conservative party")
```
Scatterplots
specifying our y limits
```{r}
CES_data %>%
  ggplot() +
  geom_point(aes( x = cps19_age,
                  y = cps19_Q_TotalDuration),
             position = "jitter", size = 0.5) +
  scale_y_continuous(limits = c(0, 1500)) +labs(x = "Age", y = "Seconds spent on the survey", title = "Age and time to complete survey amongst attentive survey takers")
```

