---
title: "Live_coding"
output: html_document
author: Julia Gallucci
date: "2023-06-03"
---
Class 3
Lesson 03

Question from class 2
When would you use an integer over a double?

In situations where you dont require a decimal, you might choose to use an integer over a double. 
By using integers, you can save memory since they typically occupy less space than doubles. 
This can be beneficial when working with large datasets or in scenarios where memory optimization is crucial. Also, mathemeatic operations on integers are generally faster than on doubles.

```{r}
age_int <- as.integer(1:100)
typeof(age_int)

age_double <- as.double(1:100)
typeof(age_double)

object.size(age_int)
object.size(age_double)
```

Question from class 2: 
str_view is likely useful for looking at nucleotide/protein sequences

```{r}
#protr package is an R package specifically designed for protein sequence analysis.
library("protr")
ls("package:protr") # list all the function
?protr
?readFASTA #FASTA format is a commonly used file format for storing biological sequences, including protein and nucleotide sequences.

P00750 <- readFASTA(system.file("protseq/P00750.fasta",
                                package = "protr"))
P00750
stringr::str_view_all(P00750, "CGA")
stringr::str_view_all(P00750, "CG")
stringr::str_count(P00750, "CG") # 4 
```

Factors
```{r}
library(forcats)
month_levels <- month.abb
month_levels
typeof(month_levels)

months <- c("Dec", "Apr", "Jan", "Mar")
months

month_fix <- factor(months, month_levels)
month_fix
view(month_fix)
fct_recode(month_fix, "December" = "Dec")
```

Recoding a factor variable
```{r}
letters_ABC <- factor(c("A","B","C","A","B","C"))
letters_ABC
table(letters_ABC)

letters_NEWABC <- fct_recode(letters_ABC,
                             "New A" ="A",
                             "New B" = "B",
                             "New C" = "C")

letters_NEWABC
```

```{r}
today()
now()
lubridate::ymd("1998-05-18")
lubridate::mdy("May 18th, 1998")
lubridate::ymd(19980518)
lubridate::dmy("18-May-1998")

today()
as_datetime(today())
now()
as_date(now())
as_datetime(now())

datetime <- ymd_hms("2016-07-08 12:34:56")
year(datetime)
month(datetime)
mday(datetime)
yday(datetime)

wday(datetime)
hour(datetime)
minute(datetime)
second(datetime)

today() - ymd("2000-01-01")

as.duration(today() - ymd("2000-01-01"))

dseconds(120)

dminutes(60)

dhours(c(12,24))

ddays(0:7)

dweeks(4)

dyears(10)

today()+ years(2)

today() + months(1)

today()+ days(1)

today() + hours(1)
today() + minutes(1)
today() + seconds(1)
today() + hours(1) +minutes(1) + seconds(1)

ymd_hms("2023-06-03 09:47:50", tz = "America/New_York")
```

Missing data
```{r}
NA > 5
NA == 10
NA + 5
NA == NA

is.na(NA)
```

Exercises
1. Make a tibble where the vectors do not have equal length. What happens?
```{r}
mytibble <- tibble(
  A = 1:10,
  B = 1
)
mytibble
```

2. In the following tibble, extract variable B:
```{r}
mytibble <- tibble(
  A = 1:10,
  B = A * 2
)
mytibble

mytibble$B
mytibble[[2]]
mytibble[["B"]]

mytibble$A
mytibble[[1]]
mytibble[["A"]]

# [row,column]
mytibble[2,2]
mytibble[2,]
```
Try using functions paste() and paste0(). Compare them to str_c(), how do they work differently?
```{r}
paste("Hi","my", "name","is","julia")
paste0("Hi","my", "name","is","julia")
str_c("Hi","my", "name","is","julia", sep = "-")
```

Look up the function str_trim() and demonstrate application.
```{r}
str_trim("Hi my name is Julia       ")
```

Given the text "Hello, world! \\." match the sequence "\" with a regular expression
```{r}
text <- "Hello, world! \\."
str_view(text, "\\\\")
```

Given the text "x-ray" match words that start with x with a regular expression
```{r}
text <- c("x-ray", "cat", "excellent")
str_view(text,"^x")
```
Given the text "cat, hat, dog, rat" match words that contain the regular expression at
```{r}
text <- c("cat", "hat", "dog", "rat")
str_view(text, "at")
```
Given the text LETTERS (A-Z), match only those of vowels with a regular expression
```{r}
str_view(LETTERS, "[AEIOU]")
```
What does ^.*$ match?
```{r}
str_view(1:10, "^.*$")

text <- "Hello world"

str_view(text, "^.*$")
```
What would happen if we parse a string with invalid dates?
```{r}
ymd("2023-02-30")
```

Lesson 04
```{r}
library(haven)
ads_data <- haven::read_sav("ads_raw.sav") #load dataset ads_raw
```

Let's view our data!
```{r}
library(dplyr)
glimpse(ads_data) 
```

Let's manipulate our data!
```{r}
filter(ads_data, Duration__in_seconds_ < 200)
#filter our data frame ads_data so that we are only extracting rows that satisfy the condition of "duration in seconds" less than 200

arrange(ads_data, -Duration__in_seconds_) #descending order

arrange(ads_data, Duration__in_seconds_) #ascending order

#arrange our data frame ads_data, so that we are sorting our values from either increasing or decreasing duration in seconds

select(ads_data, RecordedDate)
#select from our data frame only the column RecordedDate.

select(ads_data, -Consent, -DistributionChannel)
#select from our data frame all columns EXCEPT (-) Consent and DistributionChannel columns

```

Let's use the %>% pipe function to perform multiple operations at once!
```{r}
ads_data %>%
  filter(Duration__in_seconds_ < 100) %>%
  arrange(Duration__in_seconds_) %>%
  select(RecordedDate, Duration__in_seconds_)

#from our data frame ads_data we want to 
# 1. filter so that we only extract rows with a duration in seconds of less than 100
# 2. arrange our data frame by duration in seconds in increasing order
# 3. from our data frame ads_data, we want to only select the specified columns
```

Let's create a new column in our data frame
```{r}
ads_data_new <- 
  ads_data %>%
  mutate(Birthyear_add_day =
           str_c(Birthyear, "07-01")) %>%
  mutate(Birthyear_add_day =
           as_datetime(Birthyear_add_day))


ads_data_new %>%
  mutate(age = EndDate - Birthyear_add_day) %>%
  select(EndDate, Birthyear_add_day, age)

#from ads data, create a new column named age, that takes the column end data and subtracts the column birthyear add day from it.
#selecting only the specified columns
```

Let's use summary to summarize our data
```{r}
summary(ads_data)
```

Let's use pull() to extract a specific column from our data into a numeric vector
```{r}
ads_data %>%
  pull(Duration__in_seconds_)


ads_data %>%
  pull(Duration__in_seconds_) %>%
  median(na.rm= TRUE)
#from our data frame ads_data
#we want to 
#1. extract duration in seconds and vectorize the data
#2. perform a mathemetical operation on the vectorized data to extract the median of the column

ads_data %>%
  pull(Duration__in_seconds_) %>%
  mean(na.rm = TRUE)

ads_data %>%
  pull(Duration__in_seconds_) %>%
  range(na.rm = TRUE)

ads_data %>%
  pull(Duration__in_seconds_) %>%
  var(na.rm = TRUE)

ads_data %>%
  pull(Duration__in_seconds_) %>%
  sd(na.rm = TRUE)
```

Let's use summarise() to create a new data frame that summarizes our existing data with the specified mathemetical operations
```{r}
ads_data %>%
  summarise(mean_Dur_in_S = mean(Duration__in_seconds_, 
                             na.rm = TRUE),
            sd_Dur_in_S = sd(Duration__in_seconds_, 
                             na.rm= TRUE))
```

We can group by a categorical variable prior to creating a summary table
```{r}
ads_data %>%
  group_by(Gender) %>%
  summarise(count = n(),
            mean_time = mean(Duration__in_seconds_,
                             na.rm = TRUE),
            sd_time = sd(Duration__in_seconds_,
                         na.rm = TRUE))

#given our data frame ads_data
# 1. group our data by gender (1,2,3)
# summarize our data
# 2. get a count using n() of how many instances of each category there are
# 3. get summary statistics for each of the categories (mean and sd for the column duration in seconds)


#why do we use na.rm= TRUE
test <- c(1:10,NA)
test
mean(test)
mean(test, na.rm = TRUE)
```

Import data CES data
```{r}
CES_data <- read_csv("https://raw.githubusercontent.com/UofT-DSI/04-intro_r/main/Lessons-Data/ces_2019_cleaned.csv")

range(CES_data$cps19_yob)

#create a new column yob_fixed which adds 1919 to original yob column so it's more intuitive
CES_data_updated <-
  CES_data %>%
  mutate(cps19_yob_fix = cps19_yob + 1919)

range(CES_data_updated$cps19_yob_fix)

#create a new column age
CES_data_updated <-
  CES_data_updated %>%
  mutate(age = 2019 - cps19_yob_fix)

CES_data_updated %>%
  select(cps19_yob_fix, age) #select columns to compare
```

Recode gender column factors
```{r}
library(tidyverse)
CES_data_updated$cps19_gender

CES_data_updated <-
  CES_data_updated %>%
  mutate(cps19_gender_fix = factor(cps19_gender)) %>%
  mutate(cps19_gender_fix = fct_recode(cps19_gender_fix,
                                       "1" = "M",
                                       "2" = "F",
                                       "3"= "NB"))

CES_data_updated %>%
  select(cps19_gender_fix, cps19_gender)
```

perform multiple operations on our data
1. filtering
2. arranging
3. selecting 

```{r}
CES_data_updated %>%
  filter(cps19_household > 10) %>%
  arrange(-cps19_household) %>%
  select(cps19_household)
```

Mutate using an ifelse statement
ifelse(condion, TRUE,FALSE)
```{r}
CES_data_updated <-
  CES_data_updated %>%
  mutate(cps19_household = ifelse(cps19_household > 15
                                  , NA,cps19_household))
CES_data_updated %>%
  filter(cps19_household > 10) %>%
  pull(cps19_household)
```

perform multiple operations on our data
1. filtering
2. arranging
3. pulling
```{r}
CES_data_updated %>%
  filter(cps19_income_number > 100000) %>%
  arrange(-cps19_income_number) %>%
  pull(cps19_income_number)

```

Mutate using an ifelse statement
ifelse(condion, TRUE,FALSE)
then 
1. filter
2. pull
```{r}
CES_data_updated <- CES_data_updated %>%
  mutate(cps19_income_number = ifelse(cps19_income_number>= 100000000, NA,
                  cps19_income_number))

CES_data_updated %>%
  filter(cps19_income_number > 100000) %>%
  pull(cps19_income_number)
```
