---
title: "Live coding"
author: "Julia"
date: "2023-06-10"
output: pdf_document
---

Class 6 Lesson 06 cont'd
Exercises from last class
1. Write a greeting function that says “good morning”, “good afternoon”, or “good evening”, depending on the time of day.

```{r}
library(lubridate)
greeting <- function(){
  current <- now()
  current_hour <- hour(current)
  greeting_output <-
    case_when(
      current_hour > 6 && current_hour < 12 ~ "Good morning!",
      current_hour >= 12 && current_hour < 18 ~ "Good afternoon!" ,
      TRUE ~ "Good evening!"
    )
  return(greeting_output)
}
greeting()
```

2-Simulate a dataset using a normal distribution with mean 100 and standard deviation 15 as variable X, and a quadratic transformation of X as variable Y. Graph your data.
```{r}
toy_data <- tibble(
  X = rnorm(n = 100, mean = 100, sd = 15),
  Y = X^2
)

library(ggplot2)

toy_data %>%
  ggplot(aes(x=X, y=Y)) + geom_point()
```

Lesson 07
```{r}
CES_data <- read_csv("CES_data_short.csv")
#OR
CES_data <- read_csv("https://raw.githubusercontent.com/UofT-DSI/04-intro_r/main/Lessons-Data/CES_data_short.csv")
dim(CES_data)
```

Bar graphs
```{r}
CES_data %>%
  ggplot(aes(x = cps19_votechoice)) +
  geom_bar(width = 0.5) #change width of the bar
```
Histogram
```{r}
CES_data %>%
  mutate(age = 90 - cps19_yob) %>% #create a new column age
  ggplot(aes(x = age)) + 
  geom_histogram()
```

Scatterplots
```{r}
CES_data %>%
  mutate(age = 90 - cps19_yob) %>%
  ggplot(aes(x = age , y = cps19_household)) +
  geom_point(position = "jitter") #jitter so you can see overlapping points
```

Exercises
1. Barplot
```{r}
CES_data %>%
  ggplot(aes(x=cps19_votechoice)) +
  geom_bar() + coord_flip() #coord_flip to flip the x and y axis
```

2. Histogram
```{r}
CES_data %>%
  ggplot(aes(x=cps19_yob)) +
  geom_histogram()
```

3. Scatter
```{r}
CES_data  %>%
  ggplot(aes(x=cps19_yob,y=cps19_income_number)) +
  geom_point()
```

Customize plots
Labels
```{r}
CES_data %>%
  ggplot(aes(x = cps19_gender)) +
  geom_bar() +
  labs(x = "Self-identified gender",
       y = "Count",
       title = "Gender frequencies in CES data")
#lab to create labels
```

Filling based on gender variable
```{r}
CES_data %>%
  ggplot(aes(x = cps19_age,
             fill = cps19_gender)) + #fill based on a variable
  geom_histogram()
```

Filling based on specified color
```{r}
CES_data %>%
  ggplot(aes(x = cps19_age)) +
  geom_histogram(fill = "blue") #fill based on a color
```

Color to represent variable
```{r}
CES_data %>%
  ggplot(aes(x = cps19_age,
             y = cps19_household,
             color = cps19_gender)) + #color representing gender
  geom_smooth() +
  geom_point() # line and data points; share aesthetics 
```

Color visual effect
```{r}
CES_data %>%
  ggplot(aes(x = cps19_age,
             y = cps19_household)) +
  geom_smooth(color = "red") + #color data points and line
  geom_point(color = "red")
```

Size represent variable
```{r}
CES_data %>%
  ggplot(aes(x = cps19_age,
             y = cps19_household,
             size = cps19_income_number)) + #size within aesthetics to represent variable
  geom_point()
```

Size for visual effect
```{r}
CES_data %>%
  ggplot(aes(x = cps19_age,
             y = cps19_household)) +
  geom_point(size = 5) #change data point size for visual effect; outside aesthetics 
```

Share aesthetics
```{r}
CES_data %>%
  ggplot(aes(x = cps19_age,
             y = cps19_income_number)) +
  geom_point() + #geom point and smooth share 
  geom_smooth()
```

Separate aesthetics
```{r}
CES_data %>%
  ggplot() +
  geom_point(aes(x = cps19_age,
                 y = cps19_income_number)) +
  geom_line(aes(x = cps19_age,
                y = mean(cps19_income_number, na.rm = TRUE))) 
#separate aesthetics for point and line
```

Facets
```{r}
CES_data %>%
  ggplot(aes(x = cps19_age)) + 
  geom_histogram() +
  facet_wrap(facets = "cps19_gender") #facets to separate histogram by categorical var
```

Themes
```{r}
histogram_age_by_gender<-
  CES_data %>%
  ggplot(aes(x = cps19_age)) +
  geom_histogram() + 
  facet_wrap(facets = "cps19_gender") +
  theme_bw() #black and white theme
histogram_age_by_gender
#ggsave("histogram_age_by_gender", histogram_age_by_gender)
#to save ggplot
```
Graphing CES data
```{r}
CES_data %>%
  ggplot()+
  geom_bar(aes(x = cps19_votechoice)) +
  scale_x_discrete(labels = c(
    "Liberal Party" = "Liberal",
    "Conservative Party" = "Conservative",
    "Bloc Québécois" = "Bloc",
    "Green Party" = "Green",
    "People's Party" = "PPC",
    "Another party" = "Other",
    "Don't know/ Prefer not to answer"= "DK/PNTA"
  )) 
#relable x axis 
```
```{r}
CES_data %>%
  ggplot()+
  geom_bar(aes(x = factor(cps19_votechoice,
                          levels = c("Liberal Party", "Conservative Party", "Bloc Québécois", "Green Party", "People's Party", "Another party", "Don't know/ Prefer not to answer" )), fill = cps19_gender), position = "dodge", width = 0.5) +
  scale_x_discrete(labels = c(
    "Liberal Party" = "Liberal",
    "Conservative Party" = "Conservative",
    "Bloc Québécois" = "Bloc",
    "Green Party" = "Green",
    "People's Party" = "PPC",
    "Another party" = "Other",
    "Don't know/ Prefer not to answer"= "DK/PNTA")
  ) + labs(x = "Voter's Choice", y = "Count", title = " Voter's Choice by Gender") +theme(plot.title = element_text(hjust = 0.5)) #add labels and center title
```

Histograms
```{r fig.height=2, fig.width=8}
CES_data %>% 
  ggplot() +
  geom_histogram(aes(x = cps19_party_rating_24),
                 bins = 10) +
  facet_wrap(~factor(cps19_spend_educ, levels = c(
    "Spend less", "Spend about the same as now", 
    "Spend more", "Don't know/ Prefer not to answer"
  )), nrow = 1) +
  labs(x = "Conservative Party Rating", 
       y = "Count",
       title = "Views on education spending and the Conservative Party")
```

Scatterplots
```{r}
CES_data %>%
  ggplot()+
  geom_point(aes(x = cps19_age,
                 y = cps19_Q_TotalDuration),
             position = "jitter",
             size = 0.7) +
  scale_y_continuous(limits = c(0,3600)) +
  labs(x = "Age",
       y = "Seconds spend completing the survey",
       title = "Age and time to complete survey among attentive survey-takers")
```

Exercises
1. Customize Barplot
```{r fig.height=10, fig.width=12}
CES_data %>%
  ggplot(aes(x=cps19_votechoice, fill = cps19_gender)) +
  scale_fill_manual("Gender", values = c("yellow","orange","red"))+
  geom_bar(position = "dodge") +
  labs(x = "Vote Choice",
       y = "Count",
       title = "Voters for Canadian Political Parties") +
  theme(text = element_text(size = 20)) +coord_flip()

#fill by gender, manually fill values
#add labels and change size of text
#flip x and y axis
```

 2. Histogram
```{r}
CES_data %>%
  ggplot(aes(x=cps19_yob)) +
  geom_histogram(bins = 15, position = "dodge") +
  facet_wrap(facets = "cps19_votechoice", nrow = 2)+
  labs(x = "Year of Birth",
       y = "Counts",
       title = "Age distribution of voters") +
  theme_classic()+
  theme(plot.title = element_text(hjust = 0.5))

#facet by vote choice
#add labels and center title
#add classic theme
```

3. Scatter
```{r}
CES_data  %>%
  ggplot() +
  geom_point(aes(x=cps19_yob,y=cps19_income_number, 
                 size = cps19_household , 
                 color = cps19_household), 
             position = "jitter") +
  scale_color_gradient("Household",low = "red", high = "blue")+
  scale_y_continuous(limits = c(0,250000))+
  geom_line(aes(x = cps19_yob, y= mean(cps19_income_number, na.rm = TRUE)))+
    labs(x = "Year of Birth",
         y = "Income number",
         title = "Voter Age and Income < 250,000",
         caption = "Note: Line represents mean income number") +
  theme_classic()

# color and size points by household
#add labels and title
#add classic theme
```