---
title: "in-class coding"
author: "Julia Gallucci"
date: "13/12/2023"
output: html_document
---

1. fix error
 ORDERING OF OPERATIONS
 GOOD PRACTICE - NAMING OF COLUMNS
```{r}
CES_data %>%
  group_by(cps19_gender) %>%
  summarise(mean_age = mean(cps19_age)) 
```

2. fix error
ERROR IN SPELLING OF COLUMN
```{r}
CES_data %>% filter(cps19_votechoice == "Green Party")
```

3. fix error
MISSING CLOSING BRACKET
```{r}
CES_data %>%
  mutate(cps19_fed_donate = factor(cps19_fed_donate, levels = c("Yes", "No","Don't know/ Prefer not to answer")))
```

4. fix error
MISSING COMMA
```{r}
CES_data %>%
  select(cps19_province,
         cps19_age,
         cps19_gender)
```

DATA IMPORTING
```{r}
#different packages to read in different data file types

#package::function_name()

readr::read_csv("ces_2019_cleaned.csv")
readxl::read_excel() #reading in an excel
haven::read_sav("ads_raw.sav")

#specification
read_csv("ces_2019_cleaned.csv", sep= ";")
read_csv("ces_2019_cleaned.csv",skip = 2)
read_csv("ces_2019_cleaned.csv", col_names = FALSE)


write_csv(object_name, "object_name.csv")
```
PIVOTING
```{r}
wide_data <- tibble(
  year = c(2017, 2018, 2019, 2020, 2021, 2022),
  milo = c(4,6.3,8,7.9,8.1,8.1),
  kitty = c(15.6,15.9,14,12.2,10.9,9.9)
)
wide_data

long_data <- wide_data %>%
  pivot_longer(cols = c("milo","kitty"),
               names_to ="cat",
               values_to = "weight")
long_data

long_data %>%
  pivot_wider(names_from = "cat",
              values_from = "weight")
```
JOIN
```{r}
employment <- tibble(
  year = c(1990,1991,1992,1994),
  rate = c(.05,.02,.04,.02)
)
employment

housing <- tibble(
  date = c(1991, 1992, 1993, 1994, 1995),
  rate = c(0.89,0.6,0.75,0.88,0.9)
)
housing

#inner join
employment %>%
  inner_join(housing, by= c("year" = "date"))

#left join
employment %>%
  left_join(housing, by = c("year" = "date"))

#right join
employment %>%
  right_join(housing, by = c("year" = "date"))

#full join
employment %>%
  full_join(housing, by = c("year"="date"))
```
FILTERING JOINS
```{r}
employment
housing

employment %>%
  semi_join(housing, by = c("year" = "date"))

housing %>%
  semi_join(employment, by = c("date" = "year"))

employment %>%
  anti_join(housing, by = c("year" = "date"))
```
DATA TABLES
```{r}
#install.package(data.table)
library(data.table)
data("iris")
iris
dt_iris <- as.data.table(iris)
dt_iris

tables()
sapply(dt_iris, class)
haskey(dt_iris)
setkey(dt_iris, Species)
haskey(dt_iris)

#data_table[i,j,k]
dt_iris[Petal.Width > 0.5]
dt_iris[,Species]
dt_iris[,list(Species)]
#alternatively
dt_iris[,.(Species)]

dt_iris[,sum((Petal.Length + Petal.Width) > 2)]

dt_iris[Species == "setosa", sum((Petal.Length + Petal.Width)>2)]

dt_iris[Species == "setosa", .N]

dt_iris[, .N, by = .(Species)]

dt_iris[Petal.Width > 1, .(mean(Petal.Length)), by = Species]

dt_iris[order(Species, -Petal.Width)]

dt_iris[,.N, .(Petal.Length > 4, Petal.Width > 1)]

dt_iris[, lapply(.SD,mean), by = Species]
#.SD refers to subset of data
dt_iris[, head(.SD, 2), by = Species]
```


















Exercise
1. Tidy the data below:
```{r}
data_q1 <- tibble(
  group = c("treat","control"),
  survival = c(17, 11),
  deceased = c(3,9)
)

data_q1
data_q1 <- 
  data_q1 %>%
  pivot_longer(cols = c("survival","deceased"),
               names_to = "Status",
               values_to = "Patient")
data_q1
data_q1 %>%
  pivot_wider(names_from = "Status",
              values_from = "Patient")
```

2. Join the dataset flights to the dataset airlines. What should the key(s) be? What do the different types of joins look like?
```{r}
library(nycflights13)
data("flights")
data("airlines")

glimpse(flights)
glimpse(airlines)

#common variable in both data sets is carrier - use that as our key

flights %>%
  inner_join(airlines, by = "carrier")

flights %>%
  full_join(airlines, by = "carrier")

airlines_2 <- airlines[-1,] #removed row 1 to demonstrate antijoin subsetting

flights  %>%
  anti_join(airlines_2, by = "carrier")
```


3. Using flights and data.table, group based on cases that have dep_delay < 0 and those that have arr_delay > 0 and count the number in each group. 
How many groups/rows are there? How many in each group?

```{r}
dt_flight <- as.data.table(flights)

dt_flight[, .N, .(dep_delay < 0 , arr_delay > 0)]
```
PROGRAMMING
```{r}
library(ggplot2)
library(purrr)
library(tidyverse)
data("iris")
```
FUNCTION
```{r}
vowels_in_string <- function(string) {
  identified_vowels <- str_view(string, pattern = "[a,A,e,E,i,I,o,O,u,U]")
  return(identified_vowels)
}

vowels_in_string("CAT")

sum_numbers <- function(max_number) {
  return(sum(seq(1,max_number)))
}

sum_numbers(191238109238)
```

```{r}
#for loops
for (i in 1:10){
  print(i * 5)
}

nrow(iris)


for(i in 2:nrow(iris)) {
  iris$previous_combo[i] <- iris$Sepal.Length[i] + iris$Sepal.Length[i - 1]
}
iris

for(i in c("a","b","c")){
  print(paste(i,str_to_upper(i)))
}

for (i in 1:3){
  print(now() + i)
}

#collect our outputs
outputs <- c()

for (i in 1:5){
  outputs<- c(outputs, i)
}

mean(outputs)

i = 1

#while loop
while (i <= 10){
  print(i)
  i = i + 1
}
```

if/else logic combined with for loop
```{r}

for(i in 1:nrow(iris)){
 if (iris[i,"Petal.Length"] > 6){
   print("large petal length") } 
  else if(iris[i,"Petal.Width"] > 1){
    print("large petal width")
  }
  else{
    print("small petal")
    } 
}

```

```{r}
grades <- tibble(marks = c(94,87, 73, 69))

?case_when
grades %>%
  mutate(letter = case_when(
    marks >= 80 ~ "A",
    marks >= 70 ~ "B",
    marks >= 60 ~ "C",
    TRUE ~"F"
  ))
```

PURR
```{r}
iris %>%
  map_dbl(mean)
iris

iris %>%
  map_chr(typeof)

iris %>%
  map(summary)

iris$Species <- as.character(iris$Species)

iris$Species
vowels_in_string(iris$Species)


#multiple arguments
x <- list(1,1,1)
y <- list(10,20,30)

map2(x,y, ~ .x +.y)
```



