---
title: "Live coding"
author: "Julia"
date: "2023-06-05"
output: html_document
---
Class 3
Lesson 04 cont'd
```{r}
CES_data <- read_csv("https://raw.githubusercontent.com/UofT-DSI/04-intro_r/main/Lessons-Data/ces_2019_cleaned.csv")
```

```{r}
table(CES_data$cps19_province)

CES_data %>%
  filter(cps19_province == "Ontario") %>% #filtering data so that we only select rows containing the value "Ontario"
  select(cps19_prov_gov_sat,
         cps19_prov_id,
         cps19_income_number) # selecting only the specified columns of interest
```

```{r}
table(CES_data$cps19_prov_id)

CES_data %>%
  filter(cps19_province == "Ontario") %>%
  select(cps19_prov_gov_sat,
         cps19_prov_id,
         cps19_income_number) %>%
  group_by(cps19_prov_id) %>%
  summarise(median_income = median(cps19_income_number, na.rm = TRUE),
            count = n())
```

```{r}
CES_data %>%
  filter(cps19_province == "Ontario") %>%
  select(cps19_prov_gov_sat,
         cps19_prov_id,
         cps19_income_number) %>%
  mutate(cps19_prov_id = factor(cps19_prov_id, 
         levels = c("Liberal", "Progressive Conservative","NDP","Green","Another party","None", "Don't know/prefer not to answer"))) %>%
  group_by(cps19_prov_id) %>%
  summarise(median_income = median(cps19_income_number, na.rm = TRUE),
            count = n())
```

```{r}
CES_data %>%
  filter(cps19_province == "Ontario") %>%
  select(cps19_prov_gov_sat,
         cps19_prov_id,
         cps19_income_number) %>%
  group_by(cps19_prov_gov_sat) %>%
  summarise(median_income = median(cps19_income_number, na.rm = TRUE),
            count = n())
```

```{r}
CES_data %>%
  filter(cps19_province == "Ontario") %>%
  select(cps19_prov_gov_sat,
         cps19_prov_id,
         cps19_income_number) %>%
  group_by(cps19_prov_id) %>%
  summarise(median_income = median(cps19_income_number, na.rm = TRUE) ,
            count = n()) %>%
  arrange(-median_income)
```

```{r}
CES_data %>%
  filter(cps19_province == "Ontario") %>%
  select(cps19_prov_gov_sat,
         cps19_prov_id,
         cps19_income_number) %>%
  mutate(cps19_prov_id = factor(cps19_prov_id, levels = c(
    "Liberal",
    "Progressive Conservative",
    "NDP",
    "Green",
    "Another party",
    "None",
    "Don't know/prefer not to answer"
  ))) %>%
  mutate(cps19_prov_gov_sat = factor(cps19_prov_gov_sat, levels = c(
    "Not at all satisfies",
    "Not very satisfied",
    "Fairly satisfied",
    "Very satisfied",
    "Don't know/prefer not to answer"
  ))) %>%
  group_by(cps19_prov_gov_sat, cps19_prov_id) %>%
  summarise(median_income = median(cps19_income_number, na.rm = TRUE))
```

```{r}
CES_data %>%
  filter(cps19_province == "Ontario") %>%
  select(cps19_prov_gov_sat,
         cps19_prov_id,
         cps19_income_number) %>%
  mutate(cps19_prov_id = factor(cps19_prov_id, levels = c(
    "Liberal",
    "Progressive Conservative",
    "NDP",
    "Green",
    "Another party",
    "None",
    "Don't know/prefer not to answer"
  ))) %>%
  mutate(cps19_prov_gov_sat = factor(cps19_prov_gov_sat, levels = c(
    "Not at all satisfies",
    "Not very satisfied",
    "Fairly satisfied",
    "Very satisfied",
    "Don't know/prefer not to answer"
  ))) %>%
  group_by(cps19_prov_gov_sat, cps19_prov_id) %>%
  summarise(median_income = median(cps19_income_number, na.rm = TRUE)) %>%
  spread(key = cps19_prov_gov_sat,
         value = median_income)
```

1. Filter the rows in the CES_data where survey taker is between 30 and 50 years old 

```{r}
CES_data %>%
  filter(cps19_age > 30 & cps19_age < 50) %>%
  select(cps19_age) %>%
  pull(cps19_age) %>%
  range()
```

2. Filter the rows in the CES_data where the survey taker answered the cps19_votechoice question (not NA)
```{r}
CES_data %>%
  filter(!(is.na(cps19_votechoice))) %>%
  select(cps19_votechoice)
```

3. Selecting the columns cps19_age and cps19_province from CES_data
```{r}
CES_data %>%
  select(cps19_age, cps19_province)
```

4. Select all but cps19
```{r}
CES_data %>%
  select(-cps19_province)
```

1. Create new variable in CES_data that tells us whether or not a person consumes news content (cps19_news_cons == 0 minutes OR not)
```{r}
CES_data %>%
  mutate(news_watcher = ifelse(cps19_news_cons == "0 minutes", "Not watcher", "Watcher")) %>%
  select(cps19_news_cons, news_watcher)
```

2. Modify the variable cps19_income_number so that its measured in 1000s 
```{r}
CES_data %>%
  mutate(cps19_income_number = cps19_income_number / 1000)
```

1. CES_data data; group by cps19_votechoice. find median and mean rating for Trudeau
```{r}
CES_data %>%
  select(cps19_lead_rating_23, cps19_votechoice) %>%
  group_by(cps19_votechoice) %>%
  summarise(median_rating = median(cps19_lead_rating_23, na.rm = TRUE), mean_rating = mean(cps19_lead_rating_23, na.rm = TRUE))
```

1. fix error
```{r}
CES_data %>%
  group_by(cps19_gender) %>%
  summarise(mean = mean(cps19_age)) 
```

2. fix error
```{r}
CES_data %>% filter(cps19_votechoice == "Green Party")
```

3. fix error
```{r}
CES_data %>%
  mutate(cps19_fed_donate = factor(cps19_fed_donate,
                                   levels = c("Yes",
                                              "No",
                                              "Don't know/ Prefer not to answer")))
```

4. fix error
```{r}
CES_data %>%
  select(cps19_province,
         cps19_age,
         cps19_gender)
```

Class 3
Lesson 05 

```{r}
wide_data <- tibble(
  year = c(2017, 2018, 2019, 2020, 2021, 2022),
  milo = c(4,6.3,8,7.9,8.1,8.1),
  kitty = c(15.6,15.9,14,12.2,10.9, 9.9)
)
wide_data
```

```{r}
long_data <- wide_data %>%
  pivot_longer(cols = c("milo", "kitty"),
               names_to= "cat",
               values_to = "weight")
long_data
```

```{r}
long_data %>%
  pivot_wider(names_from = "cat", 
              values_from = "weight")
```

```{r}
employment <- tibble(
  year = c(1990, 1991, 1992, 1994),
  rate = c(.05,.02,.04,.02)
)
employment
```

```{r}
housing <- tibble(
  date = c(1991, 1992, 1993, 1994, 1995),
  rate = c(.89,.6,.75,.88,.9)
)

housing
```

Inner join
```{r}
employment %>%
  inner_join(housing, by = c("year" = "date"))
```

Left join
```{r}
employment %>%
  left_join(housing, by = c("year" = "date"))
```

Right join
```{r}
employment %>%
  right_join(housing, by = c("year" = "date"))
```

Full join
```{r}
employment %>%
  full_join(housing, by = c("year" = "date"))
```

Semi-join
```{r}
employment %>%
  semi_join(housing, by= c("year" = "date"))
```
Anti-join
```{r}
employment %>%
  anti_join(housing, by = c("year" = "date"))
```

Data.tables
```{r}
library(data.table)
data("iris")
dt_iris <- as.data.table(iris)
dt_iris
tables()

sapply(dt_iris, class)
haskey(dt_iris)
setkey(dt_iris, Species)
haskey(dt_iris)
```

```{r}
dt_iris[Petal.Width > 0.5]

dt_iris[,Species]

dt_iris[,list(Species)]
```